---
title: "ST558 Final Project by Wenna Han - EDA"
format: html
editor: visual
---

## Introduction
You should have an introduction section that
• briefly describes the data and the variables you have to work with (just discuss the ones you’ll investigate use in your analysis).
• describes the purpose of your EDA and ultimate goal of modeling.

For this project, we will analyze a Diabetes Health Indicators Dataset: https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset. Specifically, the diabetes_binary_health_indicators_BRFSS2015.csv data.

## Data

Use a relative path to import the data. You likely want to convert a lot of the variables to factors with
meaningful level names.
Check on missingness, etc.

Read data and correct the data type for each variable:
```{r}
data <- read.csv("diabetes_binary_health_indicators_BRFSS2015.csv") # read data
str(data) # view data structure

# select variables that should be character type
No_character<-c(1:4, 6:14,18,19) 
# correct the data type
data [,No_character] <- lapply(data[,No_character], as.character) 
str(data) # check data structure
```

## Summarizations
You should then produce meaningful summary statistics and plots about the data you are working with
(especially as it relates to your response).
Although a best practice would be to split the data at hand into a training and testing set first, go ahead
and do your EDA on the full data.
Be sure to have a narrative about what you are exploring and what the summaries and graphs
you created say about the relationships in your data.

```{r}
#| echo: false
2 * 2
```

[Click here for the Modeling Page](Modeling.html)
